---
title: "date_season_ufo"
output: html_document
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Plot date 
My questions: plot the date and see how it looks like over the years. Is there a specific season during the years that there are more sightings? Are these more apparent at night/day? A specific location? MAYBE: What are my changes of going there and seeing them? 

NOTES: 
-The Data is from 1910-2014 (104 years)
-From the span of 104 years, only 84 years have data. That means that theres a total of 20 years of consecutive no sightings of ufo. 
- 


```{r}
library(tidyverse)
library(dplyr)
library(lubridate) #for the mdy()
library(ggplot2)

#glimpse(new_ufo)

new_ufo <- new_ufo %>% 
  mutate(date = mdy(date)) %>%   # convert chr -> Date
  arrange(date) %>%                  # oldest to latest
  na.omit()


```

When and where are UFO sightings more commonly reported? What are the time of day, time of year, and location that will maximize someones chances of witnessing an UFO? 

1) how has the number of UFO sightings have changed over the years?
- make a cutoff to 2014
```{r}
years_ufo <- new_ufo %>% 
  mutate(year = year(date)) %>%  # pull out year as numeric
  count(year, name = "n")        # n = number of rows per year

ggplot(years_ufo, aes(x = year, y = n)) +
  geom_line(color = "red", linewidth = 0.8) +
  geom_point(size = 1.2, color = "blue") +
  scale_x_continuous(breaks = seq(min(years_ufo$year), max(years_ufo$year), by = 5)) +
  scale_y_continuous(limits = c(0, NA),
                     breaks = seq(0, 7000, by = 500)) +
  labs(
    title = "UFO Sightings Over the Years",
    x = "Year",
    y = "Number of Sightings"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  


```

2) Whne are UFO sightings more common during the year?
```{r}

months_ufo <- new_ufo %>%
  mutate(
    month = month(date, label = TRUE, abbr = TRUE),
    month_num = month(date),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  count(month, season, name = "n")

ggplot(months_ufo, aes(x = month, y = n, fill = season)) +
  geom_col(color="black") +
  scale_fill_manual(values = c(
    "Winter" = "blue",
    "Spring" = "green",
    "Summer" = "red",
    "Fall"   = "yellow"
  )) +
  labs(
    title = "UFO Sightings by Month and Season",
    x     = "Month",
    y     = "Number of Sightings",
    fill  = "Season"
  ) +
  theme_minimal()
```
```{r}
season_state_ufo <- new_ufo %>%
  mutate(
    month_num = month(date),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  filter(country == "us") %>%        #focusing on the united states 
  count(season, state, name = "n") %>%
  group_by(season) %>%
  slice_max(n, n = 10)               #top 10 states per season

ggplot(season_state_ufo, aes(x = reorder(state, n), y = n, fill = season)) +
  geom_col(color="black", show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ season, scales = "free_y") +
  scale_fill_manual(values = c(
    "Winter" = "blue",
    "Spring" = "green",
    "Summer" = "red",
    "Fall"   = "yellow"
  )) +
  labs(
    title = "Top 10 States for UFO Sightings by Season in the United States",
    x     = "State",
    y     = "Number of Sightings"
  ) +
  theme_minimal()
```

```{r}



new_ufo <- new_ufo %>% 
  mutate(latitude = as.numeric(latitude) )
glimpse(new_ufo)

season_map_ufo <- new_ufo %>%
  mutate(
    month_num = month(date),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  filter(country == "us",
         between(latitude, 24, 50),       # continental US bounds
         between(longitude, -125, -66))

ggplot(season_map_ufo, aes(x = longitude, y = latitude, color = season)) +
  geom_point(alpha = 0.3, size = 0.8) +  # alpha handles overplotting
  facet_wrap(~ season) +
  scale_color_manual(values = c(
    "Winter" = "steelblue",
    "Spring" = "mediumseagreen",
    "Summer" = "tomato",
    "Fall"   = "darkorange"
  )) +
  labs(
    title = "UFO Sightings Location by Season",
    x     = "Longitude",
    y     = "Latitude",
    color = "Season"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

3) Are UFO sightings more frequestly reported at night or during the day?
```{r}

glimpse(new_ufo)

time_ufo <- new_ufo %>%
  mutate(
    hour = as.integer(str_extract(time, "^\\d+")),
    day_night = if_else(hour >= 6 & hour < 20, "Day", "Night")
  ) %>%
  count(day_night, name = "n")

ggplot(time_ufo, aes(x = day_night, y = n, fill = day_night)) +
  geom_col(color="black", width = 0.5) +
  scale_fill_manual(values = c("Day" = "yellow", "Night" = "blue")) +
  labs(
    title = "UFO Sightings: Day vs. Night",
    x     = "Time of Day",
    y     = "Number of Sightings",
    fill  = "Time of Day"
  ) +
  theme_minimal()
```

Do some UFO shapes tend to be seen for longer durations than others?
```{r}
shape_counts <- new_ufo %>%
  count(shape, name = "n") %>%
  arrange(desc(n))

shape_counts


```







