---
title: "WHERE CAN YOU FIND E.T.?"
author: "Tyler Guerrero, Emanuel Padilla, Cameron Snyder"
output: 
  revealjs::revealjs_presentation:
    theme: black          
    transition: slide      
    self_contained: false 
    center: false
    reveal_options:
      slideNumber: true
      scrollable: true  
 
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(RColorBrewer)
library(ggrepel)
library(ggthemes)
library(viridis)
library(ggridges)
library(tidyverse)
library(lubridate) #for the mdy()

ufo_data <- read.csv("ufo_data.csv")
new_ufo <- ufo_data %>% select("datetime", "city", "state", "country", "shape", "duration..seconds.", "latitude", "longitude") %>% 
  mutate(across(where(is.character), trimws)) %>% #filtering the white spaces 
  filter(if_all(everything(), ~ . != "")) %>% 
  na.omit() %>% 
  separate(col=datetime, into = c("date", "time"), sep = " ")

# Adding new data set and changing strings to numeric values for data manipulation
df <- read.csv('ufo_sighting_data.csv',sep=',',stringsAsFactors=F)
df$latitude <- as.numeric(df$latitude)
df$year <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][1],"/")[[1]][3]))
df$month <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][1],"/")[[1]][1]))
df$day <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][1],"/")[[1]][2]))
df$hour <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][2],":")[[1]][1]))
df$hour <- ifelse(df$hour==24,0,df$hour)
df$min <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][2],":")[[1]][2]))
df$month_name <- sapply(df$month, function(x) month.name[x])
df$month_name_ordered <- factor(df$month_name, levels =c(month.name))
df$DATE <- sapply(df$Date_time, function(x) strsplit(x,' ')[[1]][1])
df$DATE_2 <- as.Date(df$DATE,"%m/%d/%Y")
df$weekday <- factor(weekdays(df$DATE_2, T), levels = rev(c("Mon", "Tue", "Wed", "Thu","Fri", "Sat", "Sun")))
df$length_min <- as.numeric(df$length_of_encounter_seconds)/60

#Theme to help graphs fit better
blank_theme <- theme_fivethirtyeight()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

```



# Slide 1: Intro {data-background="ETfinger.jpg"}

A Peek Into the Life of the E.T.
[Link to UFO Sightings Around the World](https://www.kaggle.com/datasets/camnugent/ufo-sightings-around-the-world)

The data is a CSV from Kaggle based on Extraterrestial Reports from 1910-2009. It shows location, time, shapes of the extraterrestial, and brief descriptions of the event.




# Slide 2: Problem and Questions {data-background="ETfinger.jpg"}

Where should you vacation to if you want a chance at meeting E.T?
 -Is there a country that best suits this need?
 -What time of year does E.T. prefer to travel?
 -Is there a particular time of day that E.T. leaves his house?
 -Could E.T. be among us at this very moment?
 



# Slide 3: Background/Reason for Choosing Data {data-background="ETfinger.jpg"}

We always hear about controversies and tinfoil hats, but we were curious how often do those from space come to visit, if at all? Is Earth a local hotspot for E.T. and his friends? Is Coachella a galaxy-wide attraction? These questions felt fun to dive into and answer.

The following libraries were used to manipulate our data and create graphs:
(ggplot2)
(dplyr)
(gridExtra)
(grid)
(RColorBrewer)
(ggrepel)
(ggthemes)
(viridis)
(ggridges)
(tidyverse)
(lubridate)




# Slide 4: Unfiltered Data {data-background="ETfinger.jpg"}

Here's a glimpse of our dataframe.
```{r}
glimpse(df)

```

 -New columns were created to alter strings to numeric values for ease of use.




# Slide 5: Key Terms and Column Focus {data-background="ETfinger.jpg"}

-UFO Shape: Describes the Manifestation of E.T. sighting
-Country abbreviations
**au - Australia
**ca - Canada
**de - Germany
**gb - Great Britain
**us - United States


# Slide 6: Overview of Data and Where {data-background="ETfinger.jpg"}

```{r}
df %>% 
  group_by(year,country) %>% 
  summarize(count=n()) %>%
  ggplot(aes(x=year,y=count,fill=country)) + 
  geom_histogram(stat='identity',width=1,color='white',size=.25) + 
  scale_fill_brewer(palette='Paired') + geom_vline(xintercept=1990,color='black') + 
  annotate("text", x = 1975, y = 6000, label = "Internet Access\nto General Public",color='black') + 
  theme_fivethirtyeight() + 
  ggtitle('UFO Sightings Overview by Year and Location', 'NA, Australia, Canada, Germany, Great Britain, United States') + 
  theme(legend.position='right',legend.direction='vertical')

```
 - With a quick overview, we can see that most reports do not actually occur until after ~1995.
 - We believe this is most likely attributed to the lack of public, internet access before then.


# Slide 7: Worldwide View {data-background="ETfinger.jpg"}

```{r}
countries_map <- map_data("world")
world_map <- ggplot() + 
  geom_map(data = countries_map, 
           map = countries_map,aes(x = long, y = lat, map_id = region, group = group),
           fill = "white", color = "green", size = 0.1)

world_map + geom_point(data=df,aes(x=longitude,y=latitude),alpha=.5,size=.25) + theme_fivethirtyeight() + ggtitle('Locations of UFO Sightings')
```

 - Here is a broader-scope view of reports around the world.
 - As we can see, there appears to be a conglomerate of reports focused in the United States.
 - Let's take a closer look...


# slide 8: United States View {data-background="ETfinger.jpg"}

```{r}
states_map <- map_data("state")

usMap <- ggplot() + 
  geom_map(data = states_map, map = states_map,aes(x = long, y = lat, map_id = region, group = group),fill = "white", color = "green", size = 0.1) + 
  theme_fivethirtyeight()

usMap + 
  geom_point(data=filter(df,country=='us' & state.province!='hi' & state.province!='ak' & state.province!='pr'),aes(x=longitude,y=latitude),alpha=.75,size=.5) + 
  ggtitle('Location of UFO Sightings in the US')

```
 - Our initial thought was that most reports would come out of the rural midwest.
 - It appears that might not be the case. Maybe a density plot will help us understand where the focus of reports are...
 
```{r}

```

```{r}
ggplot(df, aes(x=reorder(state.province, state.province, FUN=length), fill=state.province)) + 
  stat_count() +
  theme_bw() + theme(axis.text.x = element_text(angle=50, size=8, hjust=1)) + 
  xlab("State") + ylab("Number of appearances") + ggtitle("UFO sightings by states in the USA");
```

# Top states and Seasons {data-background="ETfinger.jpg"}
```{r}
season_state_ufo <- new_ufo %>%
  mutate(
    date = parse_date_time(date, orders = c("mdy", "ymd", "dmy", "mdy HM", "ymd HM")),
    month_num = month(date),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  filter(country == "us") %>%        #focusing on the united states 
  count(season, state, name = "n") %>%
  group_by(season) %>%
  slice_max(n, n = 10)               #top 10 states per season

ggplot(season_state_ufo, aes(x = reorder(state, n), y = n, fill = season)) +
  geom_col(color="black", show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ season, scales = "free_y") +
  scale_fill_manual(values = c(
    "Winter" = "blue",
    "Spring" = "green",
    "Summer" = "red",
    "Fall"   = "yellow"
  )) +
  labs(
    title = "Top 10 States for UFO Sightings by Season in the United States",
    x = "State",
    y= "Number of Sightings"
  ) +
  theme_minimal()
```

# slide: Day or Night {data-background="ETfinger.jpg"}
```{r}

ca_summer_tod <- new_ufo %>%
  mutate(
    date = parse_date_time(date, orders = c("mdy", "ymd", "dmy")),
    hour= as.integer(substr(time, 1, 2)),  # getting the  hour from the time var string
    time_of_day = case_when(
      hour >= 6 & hour < 20~ "Daytime",
      TRUE~ "Nighttime"
    ),
    month_num   = month(date),
    season      = case_when(
      month_num %in% c(12, 1, 2) ~ "Winter",
      month_num %in% c(3, 4, 5)  ~ "Spring",
      month_num %in% c(6, 7, 8)  ~ "Summer",
      month_num %in% c(9,10,11)  ~ "Fall"
    )
  ) %>%
  filter(country == "us", state == "ca", season == "Summer") %>%
  count(time_of_day, name = "n")


ggplot(ca_summer_tod, aes(x = time_of_day, y = n, fill = time_of_day)) +
  geom_col(color = "black", show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Daytime"   = "gold",
    "Nighttime" = "navy"
  )) +
  labs(
    title = "UFO Sightings by Time of Day in California (Summer)",
    x = "Time of Day",
    y= "Number of Sightings"
  ) +
  theme_minimal()
```


# Slide 10: Additional Graphs {data-background="ETfinger.jpg"}

```{r}
df %>% group_by(hour,month_name_ordered) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=factor(hour),y=count,color=month_name_ordered,group=month_name_ordered)) + 
  geom_line() + 
  theme_fivethirtyeight() + 
  scale_color_manual(name="",values=brewer.pal(12,'Paired')) + 
  geom_point(color='black',size=.5,alpha=1) + 
  ggtitle('UFO Observations During the Day','Time Based on 24-hour Cycle')

df %>% group_by(weekday,month_name_ordered) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=weekday,y=count,color=month_name_ordered,group=month_name_ordered)) + 
  geom_line() + 
  theme_fivethirtyeight() + 
  scale_color_brewer(name="",palette='Paired') + 
  geom_point(color='black',size=.5,alpha=1) + 
  ggtitle('UFO Observations During the Week')

ggplot(df, aes(x=reorder(UFO_shape, UFO_shape, FUN=length), fill=UFO_shape)) + 
  geom_bar(show.legend=F) +
  coord_flip() +
  theme_bw() + xlab("UFO Shape") + ylab("Number of Appearances") + ggtitle("UFO Shapes Seen Worldwide");

```




# Slide 11: Duration of Encounters {data-background="ETfinger.jpg"}

```{r}
df %>% filter(length_min>1e6) %>% 
  select(-c(length_of_encounter_seconds,DATE_2,description,year, month, day ,hour, min, month_name, month_name_ordered,state.province))
```

some UFO shapes tend to be seen for longer durations than others?
```{r}
shape_counts <- new_ufo %>%
  count(shape, name = "n") %>%
  arrange(desc(n))

shape_counts


```
  
  
# summary {data-background="ETfinger.jpg"}