---
title: "WHERE CAN YOU FIND E.T.?"
author: "Tyler Guerrero, Emanuel Padilla, Cameron Snyder"
output:
  slidy_presentation:
    css: neon.css
    self_contained: false
    incremental: true
---

      
```{css, echo=FALSE}
.reveal .slides {
  overflow-y: auto !important;
}

.reveal .slides section {
  overflow-y: auto !important;
  height: 100%;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(RColorBrewer)
library(ggrepel)
library(ggthemes)
library(viridis)
library(ggridges)
library(tidyverse)
library(lubridate) #for the mdy()

ufo_data <- read.csv("ufo_data.csv")
new_ufo <- ufo_data %>% select("datetime", "city", "state", "country", "shape", "duration..seconds.", "latitude", "longitude") %>% 
  mutate(across(where(is.character), trimws)) %>% #filtering the white spaces 
  filter(if_all(everything(), ~ . != "")) %>% 
  na.omit() %>% 
  separate(col=datetime, into = c("date", "time"), sep = " ")

# Adding new data set and changing strings to numeric values for data manipulation
df <- read.csv('ufo_sighting_data.csv',sep=',',stringsAsFactors=F)
df$latitude <- as.numeric(df$latitude)

dt <- mdy_hm(df$Date_time)

df$year  <- year(dt)
df$month <- month(dt)
df$day   <- day(dt)
df$hour  <- hour(dt)
df$hour  <- ifelse(df$hour == 24, 0, df$hour)
df$min   <- minute(dt)

df$month_name <- month(dt, label = FALSE, abbr = FALSE)
df$month_name_ordered <- factor(df$month_name, levels = month.name)

df$DATE   <- date(dt)
df$DATE_2 <- as.Date(df$DATE)

df$weekday <- factor(wday(dt, label = TRUE, abbr = TRUE, week_start = 1),
                     levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

df$length_min <- as.numeric(df$length_of_encounter_seconds) / 60

#Theme to help graphs fit better
blank_theme <- theme_fivethirtyeight()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

```



# Introduction
- INTRODUCTION:
- Is it a bird, is it a plane…Superman? If you can’t name it, it might just be an unidentified flying object. Have you ever stopped to wonder why we’re so fascinated by UFOs in the first place? As humans, we’re wired for curiosity and love turning mysteries into stories, theories, and yes—full‑on tinfoil‑hat conspiracies. It’s time to lean into that curiosity and actually investigate. What are these UFOs? Is E.T. and his friends real? Are they just making a pit stop on good old oil‑rich Earth for a quick refuel? The real challenge is figuring out where these crafts tend to show up so we have the best shot at actually interacting with them. After all, who wouldn’t want to hitch a ride across the universe with E.T. and hear all the cosmic secrets?

- In this analysis, we’ll hunt for potential UFO “hotspots” on Earth—places where sightings cluster—so you could, in theory, grab some friends, camp out, and go UFO‑watching in high‑probability zones. Our main UFO dataset contains descriptive witness reports dating back to 1910, obtained from Kaggle and originally compiled by the National UFO Reporting Center. We also incorporate an additional UFO dataset; together, these sources (ufo1 and ufo2) give us a richer foundation for exploring where and when unexplained sightings light up our sky.

- QUESTIONS TO ANSWER:
- Where should you vacation to if you want a chance at meeting E.T?
  - Is there a country that best suits this need?
  - What time of year does E.T. prefer to travel?
  - Is there a particular time of day that E.T. leaves his house?
  - Could E.T. be among us at this very moment?

- TOOLS USED:
  ```{r}
  #Data Manipulation:
  
  library(dplyr) # Used for data wrangling
  library(tidyverse) # Used to analyze data 
  library(lubridate) # Used to manipulate time format
  
  #Data Visualization & Aesthetics:
  
  library(ggplot2) # Used for plotting
  library(ggridges) # Used for ridge plots
  library(RColorBrewer) # Used for color palettes
  library(viridis) # Used for colorblind-safe palettes
  library(ggrepel) #Used for non-overlapping labels
  library(ggthemes) # Used for custom themes
  library(grid) # Used for low-level layout
  library(gridExtra) # Used for arranging plots
  
  ```
  
# Data Preparation
- source and cleaning:

# Data:
- Display and Summary:

# Exploratory Data Analysis
- Discovery
- plots/graphs/tables
- explanation

# Summary
- problem and insights
- implication and limitations
- formatting and other requirements

# Intro

A Peek Into the Life of the E.T.
[Link to UFO Sightings Around the World](https://www.kaggle.com/datasets/camnugent/ufo-sightings-around-the-world)

The data is a CSV from Kaggle based on Extraterrestial Reports from 1910-2009. It shows location, time, shapes of the extraterrestial, and brief descriptions of the event.




# Problem and Questions 

Where should you vacation to if you want a chance at meeting E.T?

 - Is there a country that best suits this need?
 - What time of year does E.T. prefer to travel?
 - Is there a particular time of day that E.T. leaves his house?
 - Could E.T. be among us at this very moment?
 



# Background/Reason for Choosing Data 

We always hear about controversies and tinfoil hats, but we were curious how often do those from space come to visit, if at all? Is Earth a local hotspot for E.T. and his friends? Is Coachella a galaxy-wide attraction? These questions felt fun to dive into and answer.

The following libraries were used to manipulate our data and create graphs:
(ggplot2)
(dplyr)
(gridExtra)
(grid)
(RColorBrewer)
(ggrepel)
(ggthemes)
(viridis)
(ggridges)
(tidyverse)
(lubridate)



# Unfiltered Data 

Here's a glimpse of our dataframe.

```{r, echo=FALSE}
glimpse(df)

```

 -New columns were created to alter strings to numeric values for ease of use.




# Key Terms and Column Focus 

- UFO Shape: Describes the Manifestation of E.T. sighting
- Country abbreviations
  - au - Australia
  - ca - Canada
  - de - Germany
  - gb - Great Britain
  - us - United States


# Overview of Data and Where 

```{r echo=FALSE, warning=FALSE}
df %>% 
  group_by(year,country) %>% 
  summarize(count=n()) %>%
  ggplot(aes(x=year,y=count,fill=country)) + 
  geom_histogram(stat='identity',width=1,color='white',size=.25) + 
  scale_fill_brewer(palette='Paired') + geom_vline(xintercept=1990,color='black') + 
  annotate("text", x = 1975, y = 6000, label = "Internet Access\nto General Public",color='black') + 
  theme_fivethirtyeight() + 
  ggtitle('UFO Sightings Overview by Year and Location', 'NA, Australia, Canada, Germany, Great Britain, United States') + 
  theme(legend.position='right',legend.direction='vertical')

```

 - With a quick overview, we can see that most reports do not actually occur until after ~1995.
 - We believe this is most likely attributed to the lack of public, internet access before then.


# Worldwide View

```{r echo=FALSE, warning=FALSE}
countries_map <- map_data("world")
world_map <- ggplot() + 
  geom_map(data = countries_map, 
           map = countries_map,aes(x = long, y = lat, map_id = region, group = group),
           fill = "white", color = "green", size = 0.1)

world_map + geom_point(data=df,aes(x=longitude,y=latitude),alpha=.5,size=.25) + theme_fivethirtyeight() + ggtitle('Locations of UFO Sightings')
```

 - Here is a broader-scope view of reports around the world.
 - As we can see, there appears to be a conglomerate of reports focused in the United States.
 - Let's take a closer look...


# United States View 

```{r echo=FALSE, warning=FALSE}
states_map <- map_data("state")

usMap <- ggplot() + 
  geom_map(data = states_map, map = states_map,aes(x = long, y = lat, map_id = region, group = group),fill = "white", color = "green", size = 0.1) + 
  theme_fivethirtyeight()

usMap + 
  geom_point(data=filter(df,country=='us' & state.province!='hi' & state.province!='ak' & state.province!='pr'),aes(x=longitude,y=latitude),alpha=.75,size=.5) + 
  ggtitle('Location of UFO Sightings in the US')

```
 - Our initial thought was that most reports would come out of the rural midwest.
 - It appears that might not be the case. Maybe a density plot will help us understand where the focus of reports are...
 
```{r warning=FALSE}

```

```{r echo=FALSE, warning=FALSE}
ggplot(df, aes(x=reorder(state.province, state.province, FUN=length), fill=state.province)) + 
  stat_count() +
  theme_bw() + theme(axis.text.x = element_text(angle=50, size=8, hjust=1)) + 
  xlab("State") + ylab("Number of appearances") + ggtitle("UFO sightings by states in the USA");
```

# Top states and Seasons 
```{r echo=FALSE, warning=FALSE}
season_state_ufo <- new_ufo %>%
  mutate(
    date = parse_date_time(date, orders = c("mdy", "ymd", "dmy", "mdy HM", "ymd HM")),
    month_num = month(date),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  filter(country == "us") %>%        #focusing on the united states 
  count(season, state, name = "n") %>%
  group_by(season) %>%
  slice_max(n, n = 10)               #top 10 states per season

ggplot(season_state_ufo, aes(x = reorder(state, n), y = n, fill = season)) +
  geom_col(color="black", show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ season, scales = "free_y") +
  scale_fill_manual(values = c(
    "Winter" = "blue",
    "Spring" = "green",
    "Summer" = "red",
    "Fall"   = "yellow"
  )) +
  labs(
    title = "Top 10 States for UFO Sightings by Season in the United States",
    x = "State",
    y= "Number of Sightings"
  ) +
  theme_minimal()
```

# Day or Night 
```{r echo=FALSE, warning=FALSE}

ca_summer_tod <- new_ufo %>%
  mutate(
    date = parse_date_time(date, orders = c("mdy", "ymd", "dmy")),
    hour= as.integer(substr(time, 1, 2)),  # getting the  hour from the time var string
    time_of_day = case_when(
      hour >= 6 & hour < 20~ "Daytime",
      TRUE~ "Nighttime"
    ),
    month_num   = month(date),
    season      = case_when(
      month_num %in% c(12, 1, 2) ~ "Winter",
      month_num %in% c(3, 4, 5)  ~ "Spring",
      month_num %in% c(6, 7, 8)  ~ "Summer",
      month_num %in% c(9,10,11)  ~ "Fall"
    )
  ) %>%
  filter(country == "us", state == "ca", season == "Summer") %>%
  count(time_of_day, name = "n")


ggplot(ca_summer_tod, aes(x = time_of_day, y = n, fill = time_of_day)) +
  geom_col(color = "black", show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Daytime"   = "gold",
    "Nighttime" = "navy"
  )) +
  labs(
    title = "UFO Sightings by Time of Day in California (Summer)",
    x = "Time of Day",
    y= "Number of Sightings"
  ) +
  theme_minimal()
```


# Additional Graphs 

```{r echo=FALSE, message=FALSE, warning=FALSE}
df %>% group_by(hour,month_name_ordered) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=factor(hour),y=count,color=month_name_ordered,group=month_name_ordered)) + 
  geom_line() + 
  theme_fivethirtyeight() + 
  scale_color_manual(name="",values=brewer.pal(12,'Paired')) + 
  geom_point(color='black',size=.5,alpha=1) + 
  ggtitle('UFO Observations During the Day','Time Based on 24-hour Cycle')

df %>% filter(state.province == "ca") %>% 
  group_by(hour,month_name_ordered) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=factor(hour),y=count,color=month_name_ordered,group=month_name_ordered)) + 
  geom_line() + 
  theme_fivethirtyeight() + 
  scale_color_manual(name="",values=brewer.pal(12,'Paired')) + 
  geom_point(color='black',size=.5,alpha=1) + 
  ggtitle('UFO Observations During the Day','Time Based on 24-hour Cycle')

df %>% filter(state.province == "fl") %>% 
  group_by(hour,month_name_ordered) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=factor(hour),y=count,color=month_name_ordered,group=month_name_ordered)) + 
  geom_line() + 
  theme_fivethirtyeight() + 
  scale_color_manual(name="",values=brewer.pal(12,'Paired')) + 
  geom_point(color='black',size=.5,alpha=1) + 
  ggtitle('UFO Observations During the Day','Time Based on 24-hour Cycle')

df %>% filter(state.province == "nd") %>% 
  group_by(hour,month_name_ordered) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=factor(hour),y=count,color=month_name_ordered,group=month_name_ordered)) + 
  geom_line() + 
  theme_fivethirtyeight() + 
  scale_color_manual(name="",values=brewer.pal(12,'Paired')) + 
  geom_point(color='black',size=.5,alpha=1) + 
  ggtitle('UFO Observations During the Day','Time Based on 24-hour Cycle')
```


```{r}
df %>% group_by(weekday,month_name_ordered) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=weekday,y=count,color=month_name_ordered,group=month_name_ordered)) + 
  geom_line() + 
  theme_fivethirtyeight() + 
  scale_color_brewer(name="",palette='Paired') + 
  geom_point(color='black',size=.5,alpha=1) + 
  ggtitle('UFO Observations During the Week')

ggplot(df, aes(x=reorder(UFO_shape, UFO_shape, FUN=length), fill=UFO_shape)) + 
  geom_bar(show.legend=F) +
  coord_flip() +
  theme_bw() + xlab("UFO Shape") + ylab("Number of Appearances") + ggtitle("UFO Shapes Seen Worldwide");

```




# Duration of Encounters 

```{r echo=FALSE, warning=FALSE}
df %>% filter(length_min>1e6) %>% 
  select(-c(length_of_encounter_seconds,DATE_2,description,year, month, day ,hour, min, month_name, month_name_ordered,state.province))
```

some UFO shapes tend to be seen for longer durations than others?
```{r echo=FALSE, warning=FALSE}
shape_counts <- new_ufo %>%
  count(shape, name = "n") %>%
  arrange(desc(n))

shape_counts


```
  