---
title: "A Peek Into the Life of E.T.?"
author: "Tyler Guerrero, Emanuel Padilla, Cameron Snyder"
output:
  slidy_presentation:
    css: neon.css
    self_contained: false
    incremental: true
---

      
```{css, echo=FALSE}
.reveal .slides {
  overflow-y: auto !important;
}

.reveal .slides section {
  overflow-y: auto !important;
  height: 100%;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(RColorBrewer)
library(ggrepel)
library(ggthemes)
library(viridis)
library(ggridges)
library(tidyverse)
library(lubridate) #for the mdy()

#dataset 1

ufo_data <- read.csv("ufo_data.csv")
new_ufo <- ufo_data %>% select("datetime", "city", "state", "country", "shape", "duration..seconds.", "latitude", "longitude") %>% 
  mutate(across(where(is.character), trimws)) %>% #filtering the white spaces 
  filter(if_all(everything(), ~ . != "")) %>% 
  na.omit() %>% 
  separate(col=datetime, into = c("date", "time"), sep = " ")

# Adding new data set and changing strings to numeric values for data manipulation
ufo_sighting_data <- read.csv('ufo_sighting_data.csv',sep=',',stringsAsFactors=F)
df <- ufo_sighting_data %>%
  mutate(country = case_when(
    is.na(country) ~ "none",
    country == ""  ~ "none",
    TRUE ~ country
  ))

df$latitude <- as.numeric(df$latitude)

dt <- mdy_hm(df$Date_time)

df$year  <- year(dt)
df$month <- month(dt)
df$day   <- day(dt)
df$hour  <- hour(dt)
df$hour  <- ifelse(df$hour == 24, 0, df$hour)
df$min   <- minute(dt)

df$month_name <- month(dt, label = FALSE, abbr = FALSE)
df$month_name_ordered <- factor(df$month_name, levels = month.name)

df$DATE   <- date(dt)
df$DATE_2 <- as.Date(df$DATE)

df$weekday <- factor(wday(dt, label = TRUE, abbr = TRUE, week_start = 1),
                     levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

df$length_min <- as.numeric(df$length_of_encounter_seconds) / 60

#Theme to help graphs fit better
blank_theme <- theme_fivethirtyeight()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )
```

# Greetings, Traveler

 - The data used is a CSV from Kaggle based on Extraterrestial Reports from 1910-2009. It shows location, time, shapes of the extraterrestial, and brief descriptions of the event.
 
 - [Link to UFO Sightings Around the World](https://www.kaggle.com/datasets/camnugent/ufo-sightings-around-the-world)


# Is it a bird? Is it a plane? Superman?

 - In this analysis, we’ll hunt for potential UFO “hotspots” on Earth—places where sightings cluster—so you could, in theory, grab some friends, camp out, and go UFO‑watching in high‑probability zones. 
 
 [IMAGE HERE]


# The questions that we seek answers for:

 - Where should you vacation to if you want a chance at meeting E.T?
 - Is there a country that best suits this need?
 - What time of year does E.T. prefer to travel?
 - Is there a particular time of day that E.T. leaves his house?
 - Could E.T. be among us at this very moment?


# TOOLS USED:

  - Data Manipulation:
   - deplyr, tidyverse, and lubridate were used to clean up our data by converting strings to numeric values, as well as allowing us to manipulate the data to find answers.
  
  - Data Visualization & Aesthetics:
   - ggplot2, ggridges, RColorBrewer, viridis, ggrepel, ggthemes, grid, gridExtra all helped us to make our presentation a little nicer looking.


# Is There a Reason We Should Care?

 - Aliens are cool!
 - Wouldn't you like to find out the best vacation locale to meet up with some?
 - Also, where are these reports even originating from? Maybe you don't want to risk being abducted!
 

# Here is a brief glimpse of our data:

```{r echo=FALSE, message=FALSE, warning=FALSE}
glimpse(df)
```


# Key Terms and Column Focus 

 - Country abbreviations
  - au - Australia
  - ca - Canada
  - de - Germany
  - gb - Great Britain
  - us - United States
 - State abbreviations
  - Too many to list!
 - Lat/Lon will be critical for our locating needs on the map!
 - Time of day, day, and month will be pertinant for determining the best season to find E.T.!
 - UFO Shape: Describes the Manifestation of E.T. sighting
 
 
# Overview of Data and Where 

```{r echo=FALSE, warning=FALSE}
df %>% 
  group_by(year,country) %>% 
  summarize(count=n()) %>%
  ggplot(aes(x = year, y = count, fill = country)) + 
  geom_histogram(stat='identity',width=1,color='white',size=.25) + 
  scale_fill_brewer(palette='Paired', labels = c(
    "none" = "No Origin", 
    "au" = "Australia", 
    "ca" = "Canada", 
    "de" = "Germany", 
    "gb" = "UK", 
    "us" = "USA")) + 
  geom_vline(xintercept=1990,color='black') + 
  annotate("text", x = 1975, y = 6000, label = "Internet Access\nto General Public",color='black') + 
  theme_fivethirtyeight() + 
  ggtitle('UFO Sightings Overview by Year and Location', 'NA, Australia, Canada, Germany, Great Britain, United States') + 
  theme(legend.position='right',legend.direction='vertical')

```

 - With a quick overview, we can see that most reports do not actually occur until after ~1995.
 - We believe this is most likely attributed to the lack of public internet access before then.
 - It also looks like the US has an overwhelming number of reports!


# Worldwide View

```{r echo=FALSE, warning=FALSE}
countries_map <- map_data("world")
world_map <- ggplot() + 
  geom_map(data = countries_map, 
           map = countries_map,aes(x = long, y = lat, map_id = region, group = group),
           fill = "white", color = "green", size = 0.1)

world_map + geom_point(data=df,aes(x=longitude,y=latitude),alpha=.5,size=.25) + theme_fivethirtyeight() + ggtitle('Locations of UFO Sightings')
```

 - Here is a broader-scope view of reports around the world.
 - Let's check for density readings!
 
# Density Plot
```{r}
world_map +
  geom_density_2d_filled(
    data = df,
    aes(x = longitude, y = latitude),
    alpha = 0.6,
    contour_var = "ndensity"
  ) +
  coord_quickmap() +
  scale_fill_viridis_d(name = "Density") +
  theme_fivethirtyeight() +
  ggtitle("Density of UFO Sightings")
```

 - Just as our overview promised...
 - As we can see, there appears to be a conglomerate of reports focused in the United States.
 - Let's take a closer look...


# United States View 

```{r echo=FALSE, warning=FALSE}
states_map <- map_data("state")

usMap <- ggplot() + 
  geom_map(data = states_map, map = states_map,aes(x = long, y = lat, map_id = region, group = group),fill = "white", color = "green", size = 0.1) + 
  theme_fivethirtyeight()

usMap + 
  geom_point(data=filter(df,country=='us' & state.province!='hi' & state.province!='ak' & state.province!='pr'),aes(x=longitude,y=latitude),alpha=.75,size=.5) + 
  ggtitle('Location of UFO Sightings in the US')


```
 
 - Our initial thought was that most reports would come out of the rural midwest.
 - It appears that might not be the case. Maybe a density plot will help us understand where the focus of reports are...
 
#
```{r warning=FALSE}
usMap + 
  geom_density_2d_filled(
    data = df %>% 
      filter(country == 'us',
             !state.province %in% c('hi','ak','pr'),
             !is.na(longitude), !is.na(latitude)),
    aes(x = longitude, y = latitude),
    alpha = 0.7,
    contour_var = "ndensity"
  ) +
  scale_fill_viridis_d(name = "Density") +
  ggtitle("Density of UFO Sightings in the US")
```

```{r echo=FALSE, warning=FALSE}
df %>%
  filter(!is.na(state.province)) %>%   # remove NA upfront
  count(state.province, sort = TRUE) %>%
  slice_head(n = 6) %>%
  mutate(state_full = state.name[match(toupper(state.province), state.abb)]) %>%
  filter(!is.na(state_full)) %>%       # safeguard: drop anything not matched
  ggplot(aes(x = reorder(state_full, n), y = n, fill = state_full)) +
  geom_col() +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 50, size = 9, hjust = 1),
    legend.position = "none"
  ) +
  xlab("State") +
  ylab("Number of Reports") +
  ggtitle("Top 5 States for UFO Sightings in the USA")
```

# Top states and Seasons 
```{r echo=FALSE, warning=FALSE}
season_state_ufo <- new_ufo %>%
  mutate(
    date = parse_date_time(date, orders = c("mdy", "ymd", "dmy", "mdy HM", "ymd HM")),
    month_num = month(date),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  filter(country == "us") %>%        #focusing on the united states 
  count(season, state, name = "n") %>%
  group_by(season) %>%
  slice_max(n, n = 10)               #top 10 states per season

ggplot(season_state_ufo, aes(x = reorder(state, n), y = n, fill = season)) +
  geom_col(color="black", show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ season, scales = "free_y") +
  scale_fill_manual(values = c(
    "Winter" = "blue",
    "Spring" = "green",
    "Summer" = "red",
    "Fall"   = "yellow"
  )) +
  labs(
    title = "Top 10 States for UFO Sightings by Season in the United States",
    x = "State",
    y= "Number of Sightings"
  ) +
  theme_minimal()
```

# Day or Night 
```{r echo=FALSE, warning=FALSE}

ca_summer_tod <- new_ufo %>%
  mutate(
    date = parse_date_time(date, orders = c("mdy", "ymd", "dmy")),
    hour= as.integer(substr(time, 1, 2)),  # getting the  hour from the time var string
    time_of_day = case_when(
      hour >= 6 & hour < 20~ "Daytime",
      TRUE~ "Nighttime"
    ),
    month_num   = month(date),
    season      = case_when(
      month_num %in% c(12, 1, 2) ~ "Winter",
      month_num %in% c(3, 4, 5)  ~ "Spring",
      month_num %in% c(6, 7, 8)  ~ "Summer",
      month_num %in% c(9,10,11)  ~ "Fall"
    )
  ) %>%
  filter(country == "us", state == "ca", season == "Summer") %>%
  count(time_of_day, name = "n")


ggplot(ca_summer_tod, aes(x = time_of_day, y = n, fill = time_of_day)) +
  geom_col(color = "black", show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Daytime"   = "gold",
    "Nighttime" = "navy"
  )) +
  labs(
    title = "UFO Sightings by Time of Day in California (Summer)",
    x = "Time of Day",
    y= "Number of Sightings"
  ) +
  theme_minimal()
```


# Additional Graphs 

```{r echo=FALSE, message=FALSE, warning=FALSE}
df %>%
  filter(country == "us", !is.na(hour), !is.na(month)) %>%
  mutate(
    season = case_when(
      month %in% c(12, 1, 2)  ~ "Winter",
      month %in% c(3, 4, 5)   ~ "Spring",
      month %in% c(6, 7, 8)   ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  mutate(season = factor(season, levels = c("Winter","Spring","Summer","Fall"))) %>% 
  count(hour, season) %>%
  ggplot(aes(x = hour, y = n, color = season)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  theme_bw() +
  xlab("Hour of Day") +
  ylab("Number of Sightings") +
  ggtitle("UFO Sightings by Time of Day (US) Across Seasons") +
  labs(color = "Season")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
df %>%
  filter(
    country == "us",
    state.province %in% c("ca", "ny", "fl", "wa"),
    !is.na(hour),
    !is.na(month)
  ) %>%
  mutate(
    state = toupper(state.province),
    season = case_when(
      month %in% c(12, 1, 2)  ~ "Winter",
      month %in% c(3, 4, 5)   ~ "Spring",
      month %in% c(6, 7, 8)   ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    ),
    season = factor(season, levels = c("Winter","Spring","Summer","Fall"))
  ) %>%
  count(state, hour, season) %>%
  ggplot(aes(x = hour, y = n, color = season)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = seq(0, 23, by = 4)) +
  facet_wrap(~state, ncol = 2) +
  theme_bw() +
  labs(
    x = "Hour of Day",
    y = "Number of Sightings",
    color = "Season",
    title = "UFO Sightings by Time of Day Across Seasons",
    subtitle = "California, New York, Florida, Washington"
  ) +
  scale_color_brewer(palette = "Set2")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
df %>%
  filter(country == "us", !is.na(DATE_2)) %>%
  mutate(
    weekday = wday(DATE_2, label = TRUE, abbr = TRUE, week_start = 1)
  ) %>%
  count(weekday) %>%
  ggplot(aes(x = weekday, y = n, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  theme_bw() +
  xlab("Day of Week") +
  ylab("Number of Sightings") +
  ggtitle("UFO Sightings by Day of Week (United States)")

df %>%
  filter(
    country == "us",
    state.province %in% c("ca", "ny", "fl", "wa"),
    !is.na(DATE_2)
  ) %>%
  mutate(
    state = toupper(state.province),
    weekday = wday(DATE_2, label = TRUE, abbr = TRUE, week_start = 1)
  ) %>%
  count(state, weekday) %>%
  ggplot(aes(x = weekday, y = n, group = state)) +
  geom_line(color = "darkred", size = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~state, ncol = 2) +
  theme_bw() +
  xlab("Day of Week") +
  ylab("Number of Sightings") +
  ggtitle("UFO Sightings by Day of Week (Selected States)")
```

```{r}
ggplot(df, aes(x=reorder(UFO_shape, UFO_shape, FUN=length), fill=UFO_shape)) + 
  geom_bar(show.legend=F) +
  coord_flip() +
  theme_bw() + xlab("UFO Shape") + ylab("Number of Appearances") + ggtitle("UFO Shapes Seen Worldwide");

df %>%
  filter(country == "us") %>%
  count(UFO_shape, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(UFO_shape)

top_shapes <- df %>%
  filter(country == "us", !is.na(UFO_shape), UFO_shape != "") %>%
  count(UFO_shape, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(UFO_shape)

df %>%
  filter(
    country == "us",
    UFO_shape %in% top_shapes,
    !is.na(hour)
  ) %>%
  ggplot(aes(
    x = reorder(UFO_shape, hour, FUN = median),
    y = hour
  )) +
  geom_boxplot(fill = "steelblue", outlier.alpha = 0.3) +
  coord_flip() +
  scale_y_continuous(breaks = 0:23) +
  theme_bw() +
  xlab("UFO Shape (Top 10 Most Common)") +
  ylab("Hour of Day (0–23)") +
  ggtitle("Do UFO Shapes Appear at Different Times of Day? (US)")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df %>%
  filter(!is.na(UFO_shape), UFO_shape != "") %>%
  count(UFO_shape, sort = TRUE) %>%
  slice_head(n = 10)

```



# Duration of Encounters 

```{r echo=FALSE, warning=FALSE}
df %>% filter(length_min>1e6) %>% 
  select(-c(length_of_encounter_seconds,DATE_2,description,year, month, day ,hour, min, month_name, month_name_ordered,state.province))
```

some UFO shapes tend to be seen for longer durations than others?
```{r echo=FALSE, warning=FALSE}
shape_counts <- new_ufo %>%
  count(shape, name = "n") %>%
  arrange(desc(n))

shape_counts


```
  