---
title: "UFO_Data"
author: "Tyler Guerrero, Emanuel Padilla, Cameron Snyder"
date: "2026-04-08"
output: html_document
---

## Setting up data
#Explaining the UFO data: 
Variables to get rid off: "duration..hours.min.", "comments", "date.posted"

Variables we are using: datetime, city, state, country, shape, duration..seconds., latitude, longitude.  

The UFO data-set gives us UFO sightings information. Those include date, time, location, and shape. 

"new_ufo" is are cleaned and updated dataset. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(RColorBrewer)
library(ggrepel)
library(ggthemes)
library(viridis)
library(ggridges)
library(tidyverse)

ufo_data <- read.csv("ufo_data.csv")
new_ufo <- ufo_data %>% select("datetime", "city", "state", "country", "shape", "duration..seconds.", "latitude", "longitude") %>% 
  mutate(across(where(is.character), trimws)) %>% #filtering the white spaces 
  filter(if_all(everything(), ~ . != "")) %>% 
  na.omit() %>% 
  separate(col=datetime, into = c("date", "time"), sep = " ")

# Adding new data set and changing strings to numeric values for data manipulation
df <- read.csv('ufo_sighting_data.csv',sep=',',stringsAsFactors=F)
df$latitude <- as.numeric(df$latitude)
df$year <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][1],"/")[[1]][3]))
df$month <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][1],"/")[[1]][1]))
df$day <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][1],"/")[[1]][2]))
df$hour <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][2],":")[[1]][1]))
df$hour <- ifelse(df$hour==24,0,df$hour)
df$min <- sapply(df$Date_time, function(x) as.numeric(strsplit(strsplit(x," ")[[1]][2],":")[[1]][2]))
df$month_name <- sapply(df$month, function(x) month.name[x])
df$month_name_ordered <- factor(df$month_name, levels =c(month.name))
df$DATE <- sapply(df$Date_time, function(x) strsplit(x,' ')[[1]][1])
df$DATE_2 <- as.Date(df$DATE,"%m/%d/%Y")
df$weekday <- factor(weekdays(df$DATE_2, T), levels = rev(c("Mon", "Tue", "Wed", "Thu","Fri", "Sat", "Sun")))
df$length_min <- as.numeric(df$length_of_encounter_seconds)/60

blank_theme <- theme_fivethirtyeight()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )
```

---
A Peek Into the Life of the E.T.
[Link to Extraterrestrial Sightings Dataset](https://www.kaggle.com/datasets/camnugent/ufo-sightings-around-the-world)
Our data was collected from Kaggle
---


```{r}
df %>% 
  group_by(year,country) %>% 
  summarize(count=n()) %>%
  ggplot(aes(x=year,y=count,fill=country)) + 
  geom_histogram(stat='identity',width=1,color='white',size=.25) + 
  scale_fill_brewer(palette='Paired') + geom_vline(xintercept=1990,color='black') + 
  annotate("text", x = 1975, y = 6000, label = "Internet access\nto general public",color='black') + 
  theme_fivethirtyeight() + 
  ggtitle('UFO Sightings Overview by Year and Location', 'NA, Australia, Canada, Germany, Great Britain, United States') + 
  theme(legend.position='right',legend.direction='vertical')

```



##Plot date 
emanuel
My questions: plot the date and see how it looks like over the years. Is there a specific season during the years that there are more sightings? Are these more apparent at night/day? A specific location? MAYBE: What are my changes of going there and seeing them?  

```{r}


```

