---
title: "UFO_Data"
author: "Tyler Guerrero, Emanuel Padilla, Cameron Snyder"
date: "2026-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r}
library(tidyverse)
library(dplyr)
ufo_data <- read.csv("ufo_data.csv")

glimpse(ufo_data)


```

#Explaining the UFO data: 
Variables to get rid off: "duration..hours.min.", "comments", "date.posted"

Variables we are using: datetime, city, state, country, shape, duration..seconds., latitude, longitude.  

The UFO data-set gives us UFO sightings information. Those include date, time, location, and shape. 

"new_ufo" is are cleaned and updated dataset. 

```{r}
new_ufo <- ufo_data %>% select("datetime", "city", "state", "country", "shape", "duration..seconds.", "latitude", "longitude") %>% 
  mutate(across(where(is.character), trimws)) %>% #filtering the white spaces
  filter(if_all(everything(), ~ . != "")) %>% 
  na.omit() %>% 
  separate(col=datetime, into = c("date", "time"), sep = " ")

```

```{r}
us_counts <- ufo_data %>% filter(country == "us") %>%
  geom_bar(aes(y = state)) + 
  labs(
  title = "UFO Sightings by State",
  subtitle = "Sightings listed are only in U.S."
)

us_counts
```


```{r}
library(maps)
ufo_data$state <- tolower(ufo_data$state)  ## states need to be lowercase for linking

states_map <- map_data("state") ## this gives us the lat and long for each point of each state.
  
map_data("state")
map_plot <-  ggplot(data =  votedata %>% filter(year == 2008), aes(map_id = state)) +
    geom_map(aes(fill = demVote), map = states_map) +
    expand_limits(x = states_map$long, y = states_map$lat)

map_plot 

```

##Plot date 
emanuel
My questions: plot the date and see how it looks like over time. Is there a specific season during the years that there are more sightings? Are these more apparent at night/day? A specific location? MAYBE: What are my changes of going there and seeing them?  

```{r}


```

